---
output:
  md_document:
    variant: markdown_github
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# trimr: Response Time Trimming in R

Test. For a detailed overview of how to use *trimr*, please see the vignettes.

## Installation

A stable release of *trimr* [is available on CRAN](https://cran.r-project.org/web/packages/trimr/). To install this, use:

```{r, eval = FALSE}
install.packages("trimr")
```

You can also install the latest developmental version of *trimr*. Please note, though, that this version is undergoing testing and potentially has unidentified bugs. (If you do use this version and note a bug, [please log it as an issue](https://github.com/JimGrange/trimr/issues)). To install the developmental version, you will first need to install the *devtools* package and install *trimr* directly from GitHub by using the following commands:

```{r, eval = FALSE}
# install devtools
install.packages("devtools")

# install trimr from GitHub
devools::install_github("JimGrange/trimr")
```

## Overview

*trimr* is an R package that implements most commonly-used response time trimming methods, allowing the user to go from a raw data file to a finalised data file ready for inferential statistical analysis.

The trimming functions available in *trimr* fall broadly into three families:

1. **Absolute Value Criterion**
1. **Standard Deviation Criterion**
1. **Recursive / Moving Criterion**

The latter implements the methods first suggsted by Van Selst & Jolicoeur (1994). 

## Example
In the example below, we go from a data frame containing data from 32 participants (in total, 20,518 trials) to a trimmed data set showing the mean trimmed RT for each experimental condition & participant using the modified recursive trimming procedure of Van Selst & Jolicoeur (1994):

```{r}
# load trimr's library
library(trimr)

# load the example data that ships with trimr
data(exampleData)

# look at the top of the example raw data
head(exampleData)

# perform the trimming
trimmedData <- modifiedRecursive(data = exampleData, minRT = 150, digits = 0)

# look at the trimmedData
trimmedData
```

## Installation Instructions
To install the package from GitHub, you need the devools package:

```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
```

Then *trimr* can be directly installed:

```{r, eval = FALSE}
devtools::install_github("JimGrange/trimr")
```

## References
Van Selst, M.,  & Jolicoeur, P. (1994). A solution to the effect of sample size on outlier elimination. *Quarterly Journal of Experimental Psychology, 47 (A)*, 631–650.
