---
title: "trimr: Response Time Trimming in R"
author: "James A. Grange"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This package provides (hopefully) useful functions for performing all standard methods of response time (RT) trimming used in experimental psychology. These include:

- `absoluteRT`: Absolute trimming  using an RT value as the cut-off (e.g., trimming all RTs below 150ms and above 2,000ms).
- `absoluteSD`: Absolute trimming using standard deviation as the cut-off (e.g., trimming all RTs slower than 2SDs above the group mean). 
- `perCellSD`:
- `perParticipantSD`:
- `perCellParticipantSD`:
- `simpleRecursive`:
- `nonRecursive`:
- `modifiedRecursive`:
- `hybridRecursive`:

The package ships with some example data to explore the above functions with; see `data(exampleData)`.

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
